---
title: "Analysis of the Effect of Vitamin C on ToothGrowth in Guinea Pigs"
author: "Nassuna Catherine"
date: "September 5, 2018"
output: html_document
---

Overview:
In this part of the project the ToothGrowth data in the R datasets package is going to be analysed. 

```{r loading dataset, echo=TRUE}
data(ToothGrowth)
```

###Basic exploratory analyses

```{r, echo=TRUE}
str(ToothGrowth)
```

```{r, echo=TRUE}
dim(ToothGrowth)
```

```{r, echo=TRUE}
names(ToothGrowth)
```

```{r, echo=TRUE}
head(ToothGrowth)
```

###Basic summary of the data

Visual impression of the relationship between length of odontoblasts with dose levels of vitamin C and method of delivery.

```{r loading ggplot2,echo=FALSE}
library(ggplot2)
```

```{r,echo=TRUE}
ggplot(data = ToothGrowth, mapping = aes(supp, len), size = 1) + geom_boxplot(aes(fill=supp)) + facet_wrap(~dose) + theme_bw()
```

From the boxplot, as the dosage increases, there is an increase in the length of odontoblasts(tooth growth increases). At lower dosages(0.5 and 1), the OJ induces more tooth growth than the VC. However, at high dosage(2), the median length of odontoblasts (indicative of tooth growth) appears to be almost the same.

###Hypothesis Tests

```{r subsetting OJ and VC effect at dose 2,echo=TRUE}
VC_effect_.5 = ToothGrowth$len[ToothGrowth$supp == 'VC' & ToothGrowth$dose == .5]
OJ_effect_.5 = ToothGrowth$len[ToothGrowth$supp == 'OJ' & ToothGrowth$dose == .5]
VC_effect_1 = ToothGrowth$len[ToothGrowth$supp == 'VC' & ToothGrowth$dose == 1]
OJ_effect_1 = ToothGrowth$len[ToothGrowth$supp == 'OJ' & ToothGrowth$dose == 1]
VC_effect_2 = ToothGrowth$len[ToothGrowth$supp == 'VC' & ToothGrowth$dose == 2]
OJ_effect_2 = ToothGrowth$len[ToothGrowth$supp == 'OJ' & ToothGrowth$dose == 2]
```

####Hypothesis Testing for Tooth Growth at Dosage 2

Null Hypothesis: Mean tooth growth (OJ) - Mean tooth growth(VC) = 0
Alternate Hypothesis: Mean tooth growth (OJ) - Mean tooth growth(VC) != 0
```{r, two-sided t-test, echo=TRUE}
t.test(OJ_effect_2, VC_effect_2, alternative = "two.sided", paired=FALSE, var.equal=FALSE)
```

Therefore, we fail to reject the null hypothesis. 
Thus, Mean tooth growth (OJ) - Mean tooth growth(VC) = 0

#### Hypothesis involving dose level effect
- Null hypothesis; There is no difference in mean tooth growth when dose level increases
- Alternate hypothesis; The tooth growth at higher dose level of vitamin C is not equal to that at lower dose level of vitamin C

```{r chunk13, echo=TRUE}
rbind(
  t.test(OJ_effect_1, OJ_effect_.5, alternative = "two.sided", paired=FALSE, var.equal=FALSE)$p.value[1:1], 
  t.test(OJ_effect_2, OJ_effect_1, alternative = "two.sided", paired=FALSE, var.equal=FALSE)$p.value[1:1], 
  t.test(VC_effect_1, VC_effect_.5, alternative = "two.sided", paired=FALSE, var.equal=FALSE)$p.value[1:1], 
  t.test(VC_effect_2, VC_effect_1, alternative = "two.sided", paired=FALSE, var.equal=FALSE)$p.value[1:1])
```

The null hypothesis is rejected in all cases since the p-value is much less than alpha.

##Conclusion
- There is statistical evidence that, at dose level of 0.5 and 1, the mean tooth growth corresponding to delivery of vitamin C by orange juice is different from that when ascorbic acid is used. However, there is no difference in mean tooth growth by both methods at dose level of 2.
- The mean tooth growth effect is higher for OJ than that for VC except at dose level 2, where there is no difference. 
- For each method of delivery, the mean tooth growth at any lower dose level of vitamin C is different from that obtained at a corresponding higher dose level.
- Mean tooth growth increases as dose level increases.

### Appendix I

### Assumptions:

- The variables are independent and identically distributed

- The length of odontoblasts from the ToothGrowth data is normally distributed

- There is unequal variance in tooth growth corresponding to different dosage and method of delivery.

- The data we are dealing with is unpaired since 60 guinea pigs were used.

#### Rcode for subsetting

```{r Code for Subsetting OJ and VC effect at different dose levels, echo=TRUE}
VC_effect_.5 = ToothGrowth$len[ToothGrowth$supp == 'VC' & ToothGrowth$dose == .5]
OJ_effect_.5 = ToothGrowth$len[ToothGrowth$supp == 'OJ' & ToothGrowth$dose == .5]
VC_effect_1 = ToothGrowth$len[ToothGrowth$supp == 'VC' & ToothGrowth$dose == 1]
OJ_effect_1 = ToothGrowth$len[ToothGrowth$supp == 'OJ' & ToothGrowth$dose == 1]
VC_effect_2 = ToothGrowth$len[ToothGrowth$supp == 'VC' & ToothGrowth$dose == 2]
OJ_effect_2 = ToothGrowth$len[ToothGrowth$supp == 'OJ' & ToothGrowth$dose == 2]
```

### Appendix II

More on the summary of the Toothgrowth dataset

```{r chunk7, echo=TRUE}
summary(ToothGrowth)
```

```{r chunk9, echo=TRUE}
by(ToothGrowth$len, INDICES = list(ToothGrowth$supp, ToothGrowth$dose), summary)
```