---
title: "Exponential distribution in R and the Central Limit Theorem"
author: "Nassuna Catherine"
date: "September 5, 2018"
output: html_document
---

### Overview
  In this part of the project the exponential distribution in R will be investigated and compared with the Central Limit Theorem(CLT). The CLT states that the distribution of properly normalised averages of iid variables becomes that of a standard normal as the sample size,n increases.Simulation was used to generate the exponential distribution.
  
For an exponential distribution with lambda = 0.2,
  
```{r,echo=TRUE}
theoretical_mean<-1/0.2
theoretical_variance<-1/0.2**2
```

```{r, echo=FALSE}
paste("theoretical mean = ", round(theoretical_mean, 2))
paste("theoretical variance = ", round(theoretical_variance, 2))
```

#####Consider the distribution(expd) of 1000 random exponentials
```{r, echo=TRUE}
expd<-rexp(1000,.2)
sample_mean<-mean(expd)
sample_variance<-var(expd)
hist(expd,main = "Plot of exponential distribution, rexp(1000, .2)")
legend("topright", legend = c(paste("Sample mean =", round(sample_mean, 2)), paste("sample variance = ", round(sample_variance, 2)))) 
```

The sample mean and sample variance for 1000 random exponentials are consistent with theoretical mean and theoretical variance

#####Consider the distribution of 1000 averages of 40 exponentials (exp_avs)
```{r, echo=TRUE}
exp_avs = NULL 
for (i in 1:1000) exp_avs = c(exp_avs, mean(rexp(40, .2)))
a<-mean(exp_avs)
b<-var(exp_avs)
hist(exp_avs, main = "Distribution of 1000 averages of 40 exponentials")
abline(v=a, col="red")
abline(v=theoretical_mean, col="blue")
legend("topright", legend = c(paste("Sample mean(red) =", round(a, 2)), paste("Theoretical mean(blue) = ", (theoretical_mean)), paste("Variance =", round(b, 2))))
```

This distribution is far more Gaussian compared to the original exponential distribution. It has a mean consistent with the theoretical mean. However the variance is much smaller since the distribution is now more Gaussian, reflecting the effect of having larger n as a result of the simulation.

###### When we plot a distribution of 1000 normalized averages of 40 exponentials(mns);
```{r, echo=TRUE}
mns = NULL
for (i in 1:1000) mns = c(mns, (mean(rexp(40, .2))-5)/(5/sqrt(40)))
d<-mean(mns)
e<-var(mns)
hist(mns, prob=TRUE, density = 10, ylab = "Density", main = "Distribution of 1000 Normalized averages of 40 exponentials")
abline(v=d, col="blue")
lines(density(mns), col = "black")
legend("topright", legend = c(paste("mean(blue) =", round(d, 1)), paste("variance = ", round(e, 2))))
```

The distribution obtained approximates to a bell-shape and it is close to that of a standard normal with mean 0, and variance 1 as shown in the figure. This conforms to the Central Limit Theorem.